Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-07 tex/main.tex
more

\begin{abstract}
SMT solvers have recently been extended with techniques for finding models
in presence of universally quantified formulas in some restricted fragments.
This paper introduces a translation which reduces axioms specifying a large
class of recursive functions, including well-founded (terminating),
to universally quantified formulas for which these
techniques are applicable.
An empirical evaluation confirms that the approach improves 
the performance of existing solvers on benchmarks from two sources. 
The translation is implemented as a preprocessor in the solver CVC4.
\end{abstract}

@@

Our empirical evaluation on benchmarks from the IsaPlanner proof planner~\cite{DBLP:conf/itp/JohanssonDB10}
and the Leon verifier~\cite{blanc2013overview} provides
evidence that this translation improves the effectiveness of the SMT solvers
\cvc and \ziii for finding counterexamples to verification conditions
(Section~\ref{sec:evaluation}). The approach is implemented as a preprocessor
in \cvc (Section~\ref{sec:front-end}). When invoked with a
command-line option, the solver assumes that functions introduced using the
\texttt{define-fun-rec} command are admissible and translates them accordingly.

@@

Another useful fragment of function definitions is the class of
\emph{productive} corecursive functions. Corecursive functions are functions to
a coalgebraic datatype. These functions can call themselves forever, without
their being inconsistent. Productive corecursive functions are functions that
progressively reveal parts of their potentially infinite
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by
$\con{scons} : \ty{int} \times \ty{stream} \to \ty{stream}$,
the definition
$\{\forallf{\const{e}} x.\;\, \const{e}(x) \teq \con{scons}(x,\; \const{e}(x + 1))\}$
falls within this fragment: Each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specify the functions it defines.
It is even possible to mix recursion and corecursion in the same function
\cite{blanchette-et-al-2015-fouco}. Theorem~\ref{thm:adm} can be extended to
cover such specifications, based on the observation that an infinite unfolding
of a corecursive definition yields a uniquely specified infinite object.

Beyond totality, an admissible set can contain underspecified functions
such as $\forallf{\con{f}} x : \typeint.\;\, \con{f}( x )
\teq \con{f}( x - 1 )$ or $\forallf{\con{g}} x.\allowbreak\;\, \con{g}( x
) \teq \con{g}( x )$. More generally, we conjecture that one can ignore all
tail-recursive calls (i.e., calls that occur on the right-hand side of the
definition, possibly in some $\lite$ branch) when establishing well-foundedness
or productivity, without affecting admissibility.

@@

%%% TYPESETTING: approximated, to get centering
\newcommand{\itemx}{\itemindent5em\item}

\begin{figure}[t]
\normalsize
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv_0( t\typ{\tau},\, p )$ $=$
\\[-.8\baselineskip]
 \begin{itemize}
   \itemx[] $\mathsf{if}$ $\tau = \Bool$ $\mathsf{and}$ $t = \boolop(t_1,\ldots,t_n)$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( t'_i,\, \chi_i ) = \conv_0( t_i,\, \pol( \boolop,\, i,\, p ) )$ $\mathsf{for}$ $i = 1, \ldots, n$ $\mathsf{in}$%\kern5em$\mid$
      \itemx[] $\mathsf{let}$ $\chi = \chi_1 \wedge \cdots \wedge \chi_n$ $\mathsf{in}$
      \itemx[] $\mathsf{if}$ $p = \ppos$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) ) \wedge \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ $\mathsf{if}$ $p = \pneg$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) \vee \neg \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n),\, \chi$}
\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forallf{\con{f}} \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall a : \fargtype{\con{f}}.\; u' [ \vecfarg{\con{f}}( a ) / \vec x ],\, \ltrue$}
\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forall \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall \vec x.\; u',\, \forall \vec x.\; \chi$}
\\[-.8\baselineskip]
    \end{itemize}
   \itemx[] $\mathsf{else}$
   \begin{itemize}
     \itemx[] \return \bigtuple{$t,\; \bigwedge\, \{ \exists a : \fargtype{\con{f}}.\; \vecfarg{\con{f}}( a ) \teq \vec t \mid \con{f}( \vec t ) \in \terms( t ),\, \con{f} \in \sfundefs{\Sigma} \}$}
   \end{itemize}
 \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv( \varphi )$ $=$ $\mathsf{let}$ $( \varphi',\, \_ ) = \conv_0( \varphi,\, \ppos )$ $\mathsf{in}$ $\varphi'$
\\[-1.5\baselineskip] %% TYPESETTING
\end{enumerate}
\caption{\,Definition of translation $\conv$}
\label{fig:A}
\end{figure}

>>>>

\begin{abstract}
\noindent
SMT solvers have recently been extended with techniques for finding models
in presence of universally quantified formulas in some restricted fragments.
This paper introduces a translation which reduces axioms specifying a large
class of recursive functions, including well-founded (terminating),
to universally quantified formulas for which these
techniques are applicable.
An empirical evaluation confirms that the approach improves 
the performance of existing solvers on benchmarks from two sources. 
The translation is implemented as a preprocessor in the solver CVC4.
\end{abstract}

@@

Our empirical evaluation on benchmarks from the IsaPlanner proof planner~\cite{DBLP:conf/itp/JohanssonDB10}
and the Leon verifier~\cite{blanc2013overview} provides
evidence that this translation improves the effectiveness of the SMT solvers
\cvc and \ziii for finding counterexamples to verification conditions
(Section~\ref{sec:evaluation}). The approach is implemented as a preprocessor
in \cvc (Section~\ref{sec:front-end}). When invoked with a specific
command-line option, the solver assumes that functions introduced using the
\texttt{define-fun-rec} command are admissible and translates them accordingly.

@@

Another useful fragment of function definitions is the class of
\emph{productive} corecursive functions. Corecursive functions are functions to
a coalgebraic datatype. These functions can call themselves forever, without
their being inconsistent. Productive corecursive functions are functions that
progressively reveal parts of their potentially infinite
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by
$\con{scons} : \ty{int} \times \ty{stream} \to \ty{stream}$,
the definition
$\{\forallf{\const{e}} x.\;\, \const{e}(x) \teq \con{scons}(x,\; \const{e}(x + 1))\}$
falls within this fragment: Each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specify the functions it defines.
It is even possible to mix recursion and corecursion in the same function
\cite{blanchette-et-al-2015-fouco}. Theorem~\ref{thm:adm} can be extended to
cover such specifications, based on the observation that infinite unfolding
a corecursive definition computes a unique infinite object.

Beyond totality, an admissible set can contain underspecified functions
such as $\forallf{\con{f}} x : \typeint.\allowbreak\;\, \con{f}( x )
\teq \con{f}( x - 1 )$ or $\forallf{\con{g}} x.\allowbreak\;\, \con{g}( x
) \teq \con{g}( x )$. More generally, we conjecture that one can ignore all
tail-recursive calls (i.e., calls that occur on the right-hand side of the
definition, possibly in some $\lite$ branch) when establishing well-foundedness
or productivity, without affecting admissibility.

@@

%%% TYPESETTING: approximated, to get centering
\newcommand{\itemx}{\itemindent6.35em\item}

\begin{figure}[t]
\normalsize
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv_0( t\typ{\tau},\, p )$ $=$
\\[-.8\baselineskip]
 \begin{itemize}
   \itemx[] $\mathsf{if}$ $\tau = \Bool$ $\mathsf{and}$ $t = \boolop(t_1,\ldots,t_n)$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( t'_i,\, \chi_i ) = \conv_0( t_i,\, \pol( \boolop,\, i,\, p ) )$ $\mathsf{for}$ $i = 1, \ldots, n$ $\mathsf{in}$%\kern6.35em$\mid$
      \itemx[] $\mathsf{let}$ $\chi = \chi_1 \wedge \cdots \wedge \chi_n$ $\mathsf{in}$
      \itemx[] $\mathsf{if}$ $p = \ppos$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) ) \wedge \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ $\mathsf{if}$ $p = \pneg$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) \vee \neg \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n),\, \chi$}
\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forallf{\con{f}} \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall a : \fargtype{\con{f}}.\; u' [ \vecfarg{\con{f}}( a ) / \vec x ],\, \ltrue$}
\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forall \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall \vec x.\; u',\, \forall \vec x.\; \chi$}
\\[-.8\baselineskip]
    \end{itemize}
   \itemx[] $\mathsf{else}$
   \begin{itemize}
     \itemx[] \return \bigtuple{$t,\; \bigwedge\, \{ \exists a : \fargtype{\con{f}}.\; \vecfarg{\con{f}}( a ) \teq \vec t \mid \con{f}( \vec t ) \in \terms( t ),\, \con{f} \in \sfundefs{\Sigma} \}$}
   \end{itemize}
 \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv( \varphi )$ $=$ $\mathsf{let}$ $( \varphi',\, \_ ) = \conv_0( \varphi,\, \ppos )$ $\mathsf{in}$ $\varphi'$
%\\[-1.25\baselineskip] %% TYPESETTING
\end{enumerate}
\caption{\,Definition of translation $\conv$}
\label{fig:A}
\end{figure}
